---
title: "2020_10_05_Germination2"
author: "Julin Maloof"
date: "10/3/2020"
output: html_document
---

Continuing Germination coding from August.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(rethinking)
library(brms)
library(tidyverse)
```

```{r}
germ <- read_csv("../Dimensions/hydrothermal-all-species/data/light_round1_tall.csv") %>%
  filter(wps == 0) %>%
  select(pops, temps, total_seeds, germ, day, cumulative_germ)
germ
```


Convert to one event per row:
```{r}
one_per_row <- function(df) {
  total_seed <- max(df$total_seeds, sum(df$germ))
  newdata <- tibble(id=1:total_seed, germ=0, day=max(df$day))
  df <- df %>% filter(germ>0)
  count <- 1
  if (nrow(df) > 0) {
    for (i in 1:nrow(df)) { # we look at each row of the df where germination occured
      for (j in 1:df$germ[i]) { # now update the newdata to reflect the germiantion of each seed
        newdata$germ[count] <- 1
        newdata$day[count]=df$day[i]
        count <- count+1 # count keeps track of which individual we are at in the new data
      } # for j
    } # for i
  } # if 
  return(newdata)
}

germone <- germ %>% group_by(pops, temps) %>%
  select(-cumulative_germ) %>% # not needed in this encoding (I think...in any case would need to be recalculated)
  nest() %>%
  mutate(newdata=map(data, one_per_row)) %>%
  select(-data) %>%
  unnest(newdata)

germone
```

## Filter to STDI only
```{r}
germ.stdi <- germone %>% filter(pops=="STDI") %>% dplyr::select(-pops)
germ.stdi
```

## ZI Gamma from previous script (best model so far)

```{r}
d <- list(N=nrow(germ.stdi),
          germ=germ.stdi$germ,
          temps=as.numeric(as.factor(germ.stdi$temps)),
          day=germ.stdi$day)

stanmodel1.7a <-
  "
data{
    int<lower=1> N;  // number of observations
    int germ[N];
    vector[N] day;
    int temps[N];
}
parameters{
    real<lower=0> shape; // should set lower bound
    vector[8] a; // alpha for the gamma curve, one for each temp
    vector[8] ap; // alpha for the proportion dormant, one for each temp
}
model{
    vector[N] p;
    vector[N] mu;
    a ~ normal( 0 , .5 ); //narrow priors to overcome divergent transitions
    ap ~ normal( 0, 1.5 ); 
    shape ~ exponential(.5); 
    for (i in 1:N) {
        p[i] = ap[temps[i]];
        p[i] = inv_logit(p[i]);
    }
    for ( i in 1:N ) {
        mu[i] = a[temps[i]];
         // apply the inverse link function
        mu[i] = shape * exp(-(mu[i]));
    }
    for ( i in 1:N ) 
       if ( germ[i] == 0 ) target += log_mix(p[i], 0, gamma_lccdf(day[i] | shape, mu[i]));
    for ( i in 1:N ) 
       if ( germ[i] == 1 ) target += log1m(p[i]) + gamma_lpdf(day[i] | shape, mu[i]);
    for ( i in 1:N ) 
       if ( germ[i] == 1 ) day[i] ~ gamma( shape, mu[i] );
}
"
m1.7a <- stan(model_code=stanmodel1.7a, data=d, chains=4, cores=4, control=list(adapt_delta=.99))
``` 


```{r}
precis(m1.7a, depth = 2)
```


```{r}
post <- as.data.frame(m1.7a)

post_logavgdays <- post %>% select(shape, starts_with("a["
)) # these are the shape and the log(mean(avg time to germinate))
mu_rate <- post_logavgdays %>% 
  mutate(across(-shape, ~ shape*exp(-(.))) ) %>%
  summarize(across(everything(), mean)) %>% select(-shape) # posterior mean
mu_rate
mu_shape <- mean(post$shape)

post_logitp <- post %>% select(starts_with("ap")) #logit p dormant
mu_p <- post_logitp %>%
  summarize_all(mean) %>%
  mutate_all(inv_logit)
mu_p

posterior_coef <- tibble(shape=mu_shape, temps=as.factor(unique(germ.stdi$temps)),
                         rate=t(mu_rate),
                         p=t(mu_p))

post_plot1.7a <- expand_grid(posterior_coef, day=1:28) %>%
  mutate(prop_germ=pgamma(day, shape=mu_shape, rate=rate) * (1-p))

stdi.plot <- germ %>% filter(pops=="STDI") %>% 
  select(day, temps, cumulative_germ, total_seeds) %>%
  mutate(temps=as.factor(temps),
         prop_germ=cumulative_germ/total_seeds)

post_plot1.7a %>% 
  ggplot(aes(x=day,y=prop_germ,color=temps,group=temps)) +
  geom_line() +
  geom_point(data=stdi.plot)

```

## Temperature as continuous predictor (quadratic)

prior predictions:

```{r}
nprior <- 100
days <- 1:28

temps <- c(-1.5, 0, 1.5) # scaled and centered

priors <- tibble(
  id=1:nprior,
  a = rnorm(nprior, 0 , 1 ),
  ap = rnorm(nprior, 0, 2),
  shape = rexp(nprior, .25),
  bg = rnorm(nprior, 0, .5),
  bgq = rnorm(nprior, 0, 1),
  bp = rnorm(nprior, 0, .5),
  bpq = rnorm(nprior, 0, 2))

predictions <- priors %>% 
  nest(priors=-id) %>% 
  expand_grid(temps) %>%
  mutate(params=map2(temps, priors, function(temp, prior) {
    p <- prior$ap + prior$bp*temp + prior$bpq*temp^2
    p <- inv_logit(p)
    mu <- prior$a + prior$bg*temp + prior$bgq*temp^2
    mu <- prior$shape*exp(-mu)
    tibble(p=p, mu=mu, shape=prior$shape)
    }
  )
  )
 
predictions <- predictions %>%
  select(id, temps, params) %>%
  mutate(prop_germ=map(params, ~ pgamma(days, shape=.$shape, rate=.$mu) * (1-.$p)))

predictions <- predictions %>% 
  select(id, temps,  prop_germ) %>% 
  unnest(prop_germ) %>% 
  mutate(day=rep(days, length.out=nrow(.)))

predictions %>% ggplot(aes(x=day, y=prop_germ, group=id)) +
  geom_line(alpha=.1) + facet_grid(~temps)

```




```{r}
d <- list(N=nrow(germ.stdi),
          germ=germ.stdi$germ,
          temps=as.vector(scale(germ.stdi$temps)),
          day=germ.stdi$day)

stanmodel2 <-
  "
data{
    int<lower=1> N;  // number of observations
    int germ[N];
    vector[N] day;
    real temps[N];
}
parameters{
    real<lower=0> shape; // should set lower bound
    real a; // alpha for the gamma curve, one for each temp
    real ap; // alpha for the proportion dormant, one for each temp
    real bg; // beta for temp effect on gamma curve
    real bgq; // beta for quadratic temp effect on gamma curbe
    real bp; // beta for temp effect on proportion dormant
    real bpq; // beta for quadratic temp effect on proportion dormant
}
model{
    vector[N] p;
    vector[N] mu;
    a ~ normal( 0 , 1 ); //narrow priors to overcome divergent transitions
    ap ~ normal( 0, 2 ); 
    shape ~ exponential(.25);
    bg ~ normal(0, .5);
    bgq ~ normal(0, 1);
    bp ~ normal(0, .5);
    bpq ~ normal(0, 2);
    for (i in 1:N) {
        p[i] = ap + bp*temps[i] + bpq*(temps[i]^2);
        p[i] = inv_logit(p[i]);
    }
    for ( i in 1:N ) {
        mu[i] = a + bg*temps[i] + bgq*(temps[i]^2);
         // apply the inverse link function
        mu[i] = shape * exp(-(mu[i]));
    }
    for ( i in 1:N ) 
       if ( germ[i] == 0 ) target += log_mix(p[i], 0, gamma_lccdf(day[i] | shape, mu[i]));
    for ( i in 1:N ) 
       if ( germ[i] == 1 ) target += log1m(p[i]) + gamma_lpdf(day[i] | shape, mu[i]);
    for ( i in 1:N ) 
       if ( germ[i] == 1 ) day[i] ~ gamma( shape, mu[i] );
}
"
m2 <- stan(model_code=stanmodel2, data=d, chains=4, cores=4, control=list(adapt_delta=.99))
``` 


```{r}
precis(m2)
```


```{r}
post <- as.data.frame(m2)

predict_post <- function(temp, post, days=1:28) {
    p <- post$ap + post$bp*temp + post$bpq*temp^2
    p <- inv_logit(p)
    mu <- post$a + post$bg*temp + post$bgq*temp^2
    mu <- post$shape*exp(-mu)
    postparams <- tibble(id = 1:length(p), p=p, mu=mu, shape=post$shape) %>%
      nest(params=-id)
    prop_germ <- postparams %>% 
      mutate(prop_germ=map(params, ~ pgamma(days, shape=.$shape, rate=.$mu) * (1-.$p))) %>% 
      select(-params) %>% 
      unnest(prop_germ) %>%
      mutate(day=rep(days, length.out=nrow(.)))
}

post_predictions <- tibble(realtemp = unique(germ.stdi$temps)) %>%
  mutate(scaletemp = scale(realtemp)) %>%
  mutate(predictions = map(scaletemp, predict_post, post)) %>% unnest(predictions)

post_predictions_summary <- post_predictions %>%
  group_by(realtemp, day) %>%
  summarize(low89=HPDI(prop_germ)[1], 
            high90=HPDI(prop_germ)[2],
            prop_germ=mean(prop_germ)) %>%
  mutate(temps=as.factor(realtemp))

stdi.plot <- germ %>% filter(pops=="STDI") %>% 
  select(day, temps, cumulative_germ, total_seeds) %>%
  mutate(temps=as.factor(temps),
         prop_germ=cumulative_germ/total_seeds)

post_predictions_summary %>% 
  ggplot(aes(x=day,y=prop_germ,color=temps,group=temps)) +
  geom_line() +
  geom_point(data=stdi.plot)

```

Not sure if there is some bug in my code?

But seems not
```{r}
inv_logit(-.58+-.204*-.02+(-0.204)^2*2.19)
```

centering may be making it hard for the quadratic to work

### scaled but not centered

prior predictions:

```{r}
nprior <- 100
days <- 1:28
scaledtemp <- scale(unique(germ.stdi$temps), center=FALSE)
temps <- c(min(scaledtemp), mean(scaledtemp), max(scaledtemp)) # scaled and centered

priors <- tibble(
  id=1:nprior,
  a = rnorm(nprior, 0 , 2 ),
  ap = rnorm(nprior, 0, 2),
  shape = rexp(nprior, .25),
  bg = rnorm(nprior, 0, 3),
  bgq = rnorm(nprior, 0, 2),
  bp = rnorm(nprior, 0, 3),
  bpq = rnorm(nprior, 0, 2))

predictions <- priors %>% 
  nest(priors=-id) %>% 
  expand_grid(temps) %>%
  mutate(params=map2(temps, priors, function(temp, prior) {
    p <- prior$ap + prior$bp*temp + prior$bpq*temp^2
    p <- inv_logit(p)
    mu <- prior$a + prior$bg*temp + prior$bgq*temp^2
    mu <- prior$shape*exp(-mu)
    tibble(p=p, mu=mu, shape=prior$shape)
    }
  )
  )
 
predictions <- predictions %>%
  select(id, temps, params) %>%
  mutate(prop_germ=map(params, ~ pgamma(days, shape=.$shape, rate=.$mu) * (1-.$p)))

predictions <- predictions %>% 
  select(id, temps,  prop_germ) %>% 
  unnest(prop_germ) %>% 
  mutate(day=rep(days, length.out=nrow(.)))

predictions %>% ggplot(aes(x=day, y=prop_germ, group=id)) +
  geom_line(alpha=.1) + facet_grid(~temps)

```




```{r}
d <- list(N=nrow(germ.stdi),
          germ=germ.stdi$germ,
          temps=as.vector(scale(germ.stdi$temps, center=FALSE)),
          day=germ.stdi$day)

stanmodel2.1 <-
  "
data{
    int<lower=1> N;  // number of observations
    int germ[N];
    vector[N] day;
    real temps[N];
}
parameters{
    real<lower=0> shape; // should set lower bound
    real a; // alpha for the gamma curve, one for each temp
    real ap; // alpha for the proportion dormant, one for each temp
    real bg; // beta for temp effect on gamma curve
    real bgq; // beta for quadratic temp effect on gamma curbe
    real bp; // beta for temp effect on proportion dormant
    real bpq; // beta for quadratic temp effect on proportion dormant
}
model{
    vector[N] p;
    vector[N] mu;
    a ~ normal( 0 , 5 ); //narrow priors to overcome divergent transitions
    ap ~ normal( 0, 5 ); 
    shape ~ exponential(.25);
    bg ~ normal(0, 10);
    bgq ~ normal(0, 5);
    bp ~ normal(0, 10);
    bpq ~ normal(0, 5);
    for (i in 1:N) {
        p[i] = ap + bp*temps[i] + bpq*(temps[i]^2);
        p[i] = inv_logit(p[i]);
    }
    for ( i in 1:N ) {
        mu[i] = a + bg*temps[i] + bgq*(temps[i]^2);
         // apply the inverse link function
        mu[i] = shape * exp(-(mu[i]));
    }
    for ( i in 1:N ) 
       if ( germ[i] == 0 ) target += log_mix(p[i], 0, gamma_lccdf(day[i] | shape, mu[i]));
    for ( i in 1:N ) 
       if ( germ[i] == 1 ) target += log1m(p[i]) + gamma_lpdf(day[i] | shape, mu[i]);
    for ( i in 1:N ) 
       if ( germ[i] == 1 ) day[i] ~ gamma( shape, mu[i] );
}
"
m2.1 <- stan(model_code=stanmodel2.1, data=d, chains=4, cores=4)
``` 


```{r}
precis(m2.1)
```


```{r}
post <- as.data.frame(m2.1)

predict_post <- function(temp, post, days=1:28) {
    p <- post$ap + post$bp*temp + post$bpq*temp^2
    p <- inv_logit(p)
    mu <- post$a + post$bg*temp + post$bgq*temp^2
    mu <- post$shape*exp(-mu)
    postparams <- tibble(id = 1:length(p), p=p, mu=mu, shape=post$shape) %>%
      nest(params=-id)
    prop_germ <- postparams %>% 
      mutate(prop_germ=map(params, ~ pgamma(days, shape=.$shape, rate=.$mu) * (1-.$p))) %>% 
      select(-params) %>% 
      unnest(prop_germ) %>%
      mutate(day=rep(days, length.out=nrow(.)))
}

post_predictions <- tibble(realtemp = unique(germ.stdi$temps)) %>%
  mutate(scaletemp = scale(realtemp, center=FALSE)) %>%
  mutate(predictions = map(scaletemp, predict_post, post)) %>% unnest(predictions)

post_predictions_summary <- post_predictions %>%
  group_by(realtemp, day) %>%
  summarize(low89=HPDI(prop_germ)[1], 
            high90=HPDI(prop_germ)[2],
            prop_germ=mean(prop_germ)) %>%
  mutate(temps=as.factor(realtemp))

stdi.plot <- germ %>% filter(pops=="STDI") %>% 
  select(day, temps, cumulative_germ, total_seeds) %>%
  mutate(temps=as.factor(temps),
         prop_germ=cumulative_germ/total_seeds)

post_predictions_summary %>% 
  ggplot(aes(x=day,y=prop_germ,color=temps,group=temps)) +
  geom_line() +
  geom_point(data=stdi.plot)

```

## Temperature as continuous predictor (Rongkui)

### prior predictions

```{r}
nprior <- 100
days <- 1:28

temps <- c(5, 20, 40)

priors <- tibble(
  id=1:nprior,
  a = rnorm(nprior, 0 , 2 ),
  ap = rnorm(nprior, 0, 2),
  shape = rexp(nprior, .25),
  sensg = rnorm(nprior, 0, 2),
  sensp = rnorm(nprior, 0, 2))

predictions <- priors %>% 
  nest(priors=-id) %>% 
  expand_grid(temps) %>%
  mutate(params=map2(temps, priors, function(temp, prior) {
    ttg <- exp(-(temp-20)^2 / prior$sensg)
    ttp <- exp(-(temp-20)^2 / prior$sensp)
    p <- prior$ap + ttp
    p <- inv_logit(p)
    mu <- prior$a + ttg
    mu <- prior$shape*exp(-mu)
    tibble(p=p, mu=mu, shape=prior$shape)
    }
  )
  )
 
predictions <- predictions %>%
  select(id, temps, params) %>%
  mutate(prop_germ=map(params, ~ pgamma(days, shape=.$shape, rate=.$mu) * (1-.$p)))

predictions <- predictions %>% 
  select(id, temps,  prop_germ) %>% 
  unnest(prop_germ) %>% 
  mutate(day=rep(days, length.out=nrow(.)))

predictions %>% ggplot(aes(x=day, y=prop_germ, group=id)) +
  geom_line(alpha=.1) + facet_grid(~temps)

```

```{r}
d <- list(N=nrow(germ.stdi),
          germ=germ.stdi$germ,
          temps=germ.stdi$temps,
          day=germ.stdi$day,
          opttemp=20)

stanmodel3 <-
  "
data{
    int<lower=1> N;  // number of observations
    int germ[N];
    vector[N] day;
    real temps[N];
    real opttemp;
}
parameters{
    real<lower=0> shape; // should set lower bound
    real ag; // alpha for the gamma curve
    real ap; // alpha for the proportion dormant
    real sensg; //for temp equation in gamma
    real sensp; //for temp equation in proportion
}
model{
    vector[N] p;
    vector[N] mu;
    vector[N] ttp;
    vector[N] ttg;
    ag ~ normal( 0 , .5); 
    ap ~ normal( 0, 1); 
    shape ~ exponential(.5);
    sensg ~ normal(0, 1);
    sensp ~ normal(0, 1);
    for (i in 1:N) {
        ttp[i] = exp(-(((temps[i]-opttemp)^2) / sensp)); // close to 1 at optimum; 
        p[i] = ap + ttp[i];
        p[i] = inv_logit(p[i]);
    }
    for ( i in 1:N ) {
        ttg[i] = exp(-(((temps[i]-opttemp)^2) / sensg)); // close to 1 at optimum; 
        mu[i] = ag + ttg[i];
         // apply the inverse link function
        mu[i] = shape * exp(-(mu[i]));
    }
    for ( i in 1:N ) 
       if ( germ[i] == 0 ) target += log_mix(p[i], 0, gamma_lccdf(day[i] | shape, mu[i]));
    for ( i in 1:N ) 
       if ( germ[i] == 1 ) target += log1m(p[i]) + gamma_lpdf(day[i] | shape, mu[i]);
    for ( i in 1:N ) 
       if ( germ[i] == 1 ) day[i] ~ gamma( shape, mu[i] );
}
"
m3 <- stan(model_code=stanmodel3, data=d, chains=4, cores=4, control = list(adapt_delta = 0.95))
``` 


```{r}
precis(m3)
```

```{r}
pairs(m3)
```



```{r}
post <- as.data.frame(m3)

predict_post <- function(temp, post, days=1:28) {
    p <- post$ap + post$bp*temp + post$bpq*temp^2
    p <- inv_logit(p)
    mu <- post$a + post$bg*temp + post$bgq*temp^2
    mu <- post$shape*exp(-mu)
    postparams <- tibble(id = 1:length(p), p=p, mu=mu, shape=post$shape) %>%
      nest(params=-id)
    prop_germ <- postparams %>% 
      mutate(prop_germ=map(params, ~ pgamma(days, shape=.$shape, rate=.$mu) * (1-.$p))) %>% 
      select(-params) %>% 
      unnest(prop_germ) %>%
      mutate(day=rep(days, length.out=nrow(.)))
}

post_predictions <- tibble(realtemp = unique(germ.stdi$temps)) %>%
  mutate(scaletemp = scale(realtemp, center=FALSE)) %>%
  mutate(predictions = map(scaletemp, predict_post, post)) %>% unnest(predictions)

post_predictions_summary <- post_predictions %>%
  group_by(realtemp, day) %>%
  summarize(low89=HPDI(prop_germ)[1], 
            high90=HPDI(prop_germ)[2],
            prop_germ=mean(prop_germ)) %>%
  mutate(temps=as.factor(realtemp))

stdi.plot <- germ %>% filter(pops=="STDI") %>% 
  select(day, temps, cumulative_germ, total_seeds) %>%
  mutate(temps=as.factor(temps),
         prop_germ=cumulative_germ/total_seeds)

post_predictions_summary %>% 
  ggplot(aes(x=day,y=prop_germ,color=temps,group=temps)) +
  geom_line() +
  geom_point(data=stdi.plot)

```