---
title: '2020-01-31'
author: "Julin N Maloof"
date: "1/30/2020"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, autodep = TRUE)
```

```{r}
library(tidyverse)
library(rethinking)
```

## Q1)
a) subset the data for day 35
b) create a new column "stem_length" that is the sum of epi, int1, int2, and int3
c) although flats are listed as 1-6, flats in sun and shade are separate. Create a new column "flat2" that corrects for this.

```{r}
data <- read_csv("figure4phyE.csv")
head(data)
```

```{r}
data35 <- data %>%
  filter(day==35) %>%
  mutate(stem_length=epi + int1 + int2 + int3,
         flat2=as.integer(as.factor(str_c(treatment, flat))),
         shade_i=ifelse(treatment=="sun", 0L, 1L),
         g_i= as.integer(factor(genotype,
                                   levels=c("Moneymaker", 
                                            "phyB1", 
                                            "phyB2", 
                                            "phyB1/B2", 
                                            "phyEami3", 
                                            "phyEami7")))) %>%
  select(genotype, treatment, g_i, shade_i, flat2, stem_length)
data35
```

note that this is not a balanced design:

```{r}
with(data35, table(genotype, flat2))
```


Ultimately you want to know if any of the mutants have a different length from Moneymaker, in sun or in shade, or if the response to shade differs.

## Q2) Fit some models to explore treatment, genotype, and flat.

### a) don't include flat.  Determine whether genotype, treatment, and their interaction are important predictors of stem_length


```{r}
mean(data35$stem_length)
sd(data35$stem_length)
```


```{r}
datsmall <- data35 %>% select(stem_length, g_i, shade_i)
mq2a1 <- ulam(alist(stem_length ~ dnorm(mu,sigma),
                  mu <- alpha[g_i] + b_shade*shade_i + b_i[g_i]*shade_i,
                  alpha[g_i] ~ dnorm(125,50),
                  b_shade ~ dnorm(0, 50),
                  b_i[g_i] ~  dnorm(0, 50),
                  sigma ~ dexp(1)),
            data=datsmall,
            chains=4,
            cores=4,
            log_lik = TRUE)
  
            
```

```{r}
precis(mq2a1, depth=2)
traceplot(mq2a1)
pairs(mq2a1)
extract.samples(mq2a1) %>%
  as.data.frame() %>%
  cor() %>%
  round(2)
```

Right, to have beta_shade and separate betas for each shade for each species is redundant

```{r}
datsmall <- data35 %>% select(stem_length, g_i, shade_i)
mq2a2 <- ulam(alist(stem_length ~ dnorm(mu,sigma),
                  mu <- alpha[g_i] + b_shade[g_i]*shade_i,
                  alpha[g_i] ~ dnorm(125,50),
                  b_shade[g_i] ~  dnorm(0, 50),
                  sigma ~ dexp(1)),
            data=datsmall,
            chains=4,
            cores=4,
            log_lik = TRUE)
```

```{r}
precis(mq2a2, depth = 2)
traceplot(mq2a2)
pairs(mq2a2)
extract.samples(mq2a2) %>%
  as.data.frame() %>%
  cor() %>%
  round(2)
```

This is sampled much better.

let's compare to some simpler models

Same shade response per genotype

```{r}
datsmall <- data35 %>% select(stem_length, shade_i, g_i)
mq2a3 <- ulam(alist(stem_length ~ dnorm(mu,sigma),
                  mu <- alpha[g_i] + b_shade*shade_i,
                  alpha[g_i] ~ dnorm(125,50),
                  b_shade ~ dnorm(0, 50),
                  sigma ~ dexp(1)),
            data=datsmall,
            chains=4,
            cores=4,
            log_lik = TRUE)
```

```{r}
precis(mq2a3)
traceplot(mq2a3)
pairs(mq2a3)
extract.samples(mq2a3) %>%
  as.data.frame() %>%
  cor() %>%
  round(2)
```

```{r}
compare(mq2a2, mq2a3)
```

Interesting...that would argue for the simple model (no difference in shade response between genotypes).

Any difference in genotypes at all?

```{r}
datsmall <- data35 %>% select(stem_length, shade_i)
mq2a4 <- ulam(alist(stem_length ~ dnorm(mu,sigma),
                  mu <- alpha + b_shade*shade_i,
                  alpha ~ dnorm(125,50),
                  b_shade ~ dnorm(0, 50),
                  sigma ~ dexp(1)),
            data=datsmall,
            chains=4,
            cores=4,
            log_lik = TRUE)
```

```{r}
precis(mq2a4)
traceplot(mq2a4)
pairs(mq2a4)
extract.samples(mq2a4) %>%
  as.data.frame() %>%
  cor() %>%
  round(2)
```

```{r}
compare(mq2a2, mq2a3, mq2a4)
```

OK genotype impt

b) starting with your best model from a), include flat without pooling

```{r}
datsmall <- data35 %>% select(stem_length, shade_i, g_i, flat2)
mq2b1 <- ulam(alist(stem_length ~ dnorm(mu,sigma),
                  mu <- alpha[g_i] + b_shade*shade_i + b_fl[flat2],
                  alpha[g_i] ~ dnorm(125,50),
                  b_shade ~ dnorm(0, 50),
                  b_fl[flat2] ~ dnorm(0,10),
                  sigma ~ dexp(1)),
            data=datsmall,
            chains=4,
            cores=4,
            log_lik = TRUE)
```


```{r}
precis(mq2b1, depth=2)
traceplot(mq2b1, ask=FALSE)
pairs(mq2b1)
extract.samples(mq2b1) %>%
  as.data.frame() %>%
  cor() %>%
  round(2)
```

c) starting with your best model from a), use a hierarchical model that allows partial pooling across flats


```{r}
datsmall <- data35 %>% select(stem_length, shade_i, g_i, flat2)
mq2c1 <- ulam(alist(stem_length ~ dnorm(mu,sigma),
                  mu <- alpha[g_i] + b_shade*shade_i + b_fl[flat2],
                  alpha[g_i] ~ dnorm(125,50),
                  b_shade ~ dnorm(0, 50),
                  b_fl[flat2] ~ dnorm(0,sigma_fl),
                  sigma ~ dexp(1),
                  sigma_fl ~ dcauchy(0,3)),
            data=datsmall,
            chains=4,
            cores=4,
            iter=4000,
            log_lik = TRUE)
```


```{r}
precis(mq2c1, depth=2)
traceplot(mq2c1, ask=FALSE)
pairs(mq2c1)
extract.samples(mq2c1) %>%
  as.data.frame() %>%
  cor() %>%
  round(2)
```

what is happening with flat in these models?  Are shade flats estimated to have a positive effect and sun flats negative?

First which flats are which?
```{r}
data35 %>%
  select(treatment, flat2) %>%
  unique() %>%
  arrange(flat2)
```

```{r}
flat.effs.b1 <- tibble(estimate=coef(mq2b1),
                       coef=names(coef(mq2b1))) %>%
  filter(str_detect(coef,"b_fl")) %>%
  mutate(treatment=rep(c("shade","sun"), each=6))
flat.effs.b1
flat.effs.b1 %>% group_by(treatment) %>%
  summarize(mean=mean(estimate), sem=sd(estimate)/sqrt(6))
```

```{r}
flat.effs.c1 <- tibble(estimate=coef(mq2c1),
                       coef=names(coef(mq2c1))) %>%
  filter(str_detect(coef,"b_fl")) %>%
  mutate(treatment=rep(c("shade","sun"), each=6))
flat.effs.c1
flat.effs.c1 %>% group_by(treatment) %>%
  summarize(mean=mean(estimate), sem=sd(estimate)/sqrt(6))
```


Q3) Compare the models, which is preferred?

```{r}
compare(mq2a3, mq2b1, mq2c1)
```

So we aren't really gaining anything by pooling (or even by adding flat) but not hurting much either.  keep it in.

how about model2 with genotype shade interaction?

```{r}
datsmall <- data35 %>% select(stem_length, shade_i, g_i, flat2)
mq2c2 <- ulam(alist(stem_length ~ dnorm(mu,sigma),
                  mu <- alpha[g_i] + b_shade[g_i]*shade_i + b_fl[flat2],
                  alpha[g_i] ~ dnorm(125,50),
                  b_shade[g_i] ~ dnorm(0, 50),
                  b_fl[flat2] ~ dnorm(0,sigma_fl),
                  sigma ~ dexp(1),
                  sigma_fl ~ dcauchy(0,3)),
            data=datsmall,
            chains=4,
            cores=4,
            iter=4000,
            log_lik = TRUE)
```


```{r}
precis(mq2c2, depth=2)
traceplot(mq2c2, ask=FALSE)
pairs(mq2c2)
extract.samples(mq2c2) %>%
  as.data.frame() %>%
  cor() %>%
  round(2)
```

```{r}
compare(mq2c1, mq2c2)
```

Q4) Using the hierarchical model, make posterior predictions

a) for average cluster

```{r}
post <- extract.samples(mq2c2)
names(post)
str(post)
```

```{r}
link_avg <- function(genotype, shade) {
  with(post, alpha[,genotype] + shade*b_shade[,genotype])
}
```

create a data frame to hold the results
```{r}
pred.df <- data35 %>% 
  select(-stem_length, -flat2) %>% 
  unique() %>%
  mutate(treatment=factor(treatment, levels = c("sun", "shade")) ) # so the plot order is correct
pred.df
```

get predictions from posterior for each genotype, shade combo, for the average flat
```{r}
pred.df.avg <- pred.df %>%
  mutate(average_response=map2(g_i, shade_i, link_avg),
         pet_length=map_dbl(average_response, mean),
         low.89=map_dbl(average_response, ~ HPDI(.)[1]),
         high.89=map_dbl(average_response, ~ HPDI(.)[2]))
pred.df.avg
```

plot it
```{r}
pred.df.avg %>%
  ggplot(aes(x=genotype, y=pet_length, ymin=low.89, ymax=high.89, fill=treatment)) +
  geom_col(position = "dodge") +
  geom_errorbar(width = .5, position = position_dodge(width=.9)) +
  ggtitle("prediction averaged across flat")
```


b) for same clusters.

This could probably also be done with link somehow

```{r}
link_flat <- function(genotype, shade) {
  with(post, alpha[,genotype] + 
         shade*b_shade[,genotype] +
         b_fl[, sample(ncol(b_fl), size=1)]) # pick a flat at random
}
```

get predictions from posterior for each genotype, shade combo, randomly sampling flats as we go.
```{r}
pred.df.flat <- pred.df %>%
  mutate(average_response=map2(g_i, shade_i, link_flat),
         pet_length=map_dbl(average_response, mean),
         low.89=map_dbl(average_response, ~ HPDI(.)[1]),
         high.89=map_dbl(average_response, ~ HPDI(.)[2]))
pred.df.flat
```

plot it
```{r}
pred.df.flat %>%
  ggplot(aes(x=genotype, y=pet_length, ymin=low.89, ymax=high.89, fill=treatment)) +
  geom_col(position = "dodge") +
  geom_errorbar(width = .5, position = position_dodge(width=.9)) +
  ggtitle("prediction including flat variance")
```


c) showing the "marginal" from cluster

McElreath uses the mean estimated sigma but wouldn't it make more sense to sample that from the posterior?  OTOH wouldn't it make the most since to simulate a certain # of flats? (I am not doing that yet).
```{r}
link_marg <- function(genotype, shade) {
  with(post, alpha[,genotype] + shade*b_shade[,genotype] + rnorm(length(sigma_fl), 0, sigma_fl))
}
```

get predictions from posterior for each genotype, shade combo, for a new set of flats
```{r}
pred.df.marg <- pred.df %>%
  mutate(average_response=map2(g_i, shade_i, link_marg),
         pet_length=map_dbl(average_response, mean),
         low.89=map_dbl(average_response, ~ HPDI(.)[1]),
         high.89=map_dbl(average_response, ~ HPDI(.)[2]))
pred.df.marg
```

plot it
```{r}
pred.df.marg %>%
  ggplot(aes(x=genotype, y=pet_length, ymin=low.89, ymax=high.89, fill=treatment)) +
  geom_col(position = "dodge") +
  geom_errorbar(width = .5, position = position_dodge(width=.9)) +
  ggtitle("prediction marginal to flat")
```

d) showing new clusters.

Here we should be able to plot individual plants from `pred.df.marg`

50 pulls from the posterior
```{r}
plot.matrix <- sapply(pred.df.marg$average_response, function(x) x[1:50])
colnames(plot.matrix) <- str_c(pred.df.marg$genotype, "_", pred.df.marg$treatment)
head(plot.matrix )
```

convert matrix to ggplot friendly form

```{r}
plot.tbl <- plot.matrix %>%
  as.tibble() %>%
  mutate(post.sample=1:nrow(.)) %>%
  gather(key="geno_trt", value="stem_length", -post.sample)
  
head(plot.tbl)
```


```{r}
plot.tbl %>%
  ggplot(aes(x=geno_trt,y=stem_length, group=post.sample)) +
  geom_line(alpha=0.2) +
  theme(axis.text.x = element_text(angle = 90,hjust=1))
```


Q5) Reparameterize the model to help with divergent transitions (even if there aren't any)

Q6--optional)
a) Which genotypes differ from MoneyMaker in Sun conditions?

We compare posterior estimates for alpha for Moneymaker vs the others and ask how often the difference is non-zero in a particular direction
```{r}
avg.MM.sun <- pred.df.avg %>%
  filter(genotype=="Moneymaker", treatment=="sun") %>%
  pull(average_response)

pred.df.avg %>%
  mutate(dif.from.MM=map(average_response, ~ avg.MM.sun - .))
```


b) Which genotypes differ from MoneyMaker in Shade conditions?
c) Which genotypes differ from MoneyMaker in their response to shade (difference in sun vs shade)?