---
title: "Chapter 11"
author: "Julin N Maloof"
date: "10/31/2019"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(rethinking)
library(tidyverse)
```

0E1
10E2
10E3
10M1
 
Problems 1 and 2 at https://github.com/rmcelreath/statrethinking_winter2019/blob/master/homework/week06.pdf

## 10E1
_If an event has probability 0.35, what are the log-odds of this event?_

```{r}
logit(.35)
log(.35/.65)
```

## 10E2
_If an event has log-odds 3.2, what is the probability of this event?_
```{r}
inv_logit(3.2)
exp(3.2) / (1+exp(3.2)) # algebra works!
```

## 10E3
_Suppose that a coefficient in a logistic regression has value 1.7. What does this imply about the proportional change in odds of the outcome?_

The increases the probability of the event by 70%

## 10M1
_Asexplainedinthechapter,binomialdatacanbeorganizedinaggregatedanddisaggregated forms, without any impact on inference. But the likelihood of the data does change when the data are converted between the two formats. Can you explain why?_

extra parameter blah blah

## PDF 1

_The data in data(NWOGrants) are outcomes for scientific funding applications for the Netherlands Organization for Scientific Research (NWO) from 2010–2012 (see van der Lee and Ellemers doi:10.1073/pnas.1510159112). These data have a very similar structure to the UCBAdmit data discussed in Chapter 11. I want you to consider a similar question: What are the total and indirect causal effects of gender on grant awards? Consider a mediation path (a pipe) through dis- cipline. Draw the corresponding DAG and then use one or more binomial GLMs to answer the question. What is your causal interpretation? If NWO’s goal is to equalize rates of funding between the genders, what type of intervention would be most effective?_

```{r}
g <- dagitty("dag{
  G -> A;
  G -> D;
  D -> A
}")
coordinates(g) <- list(x=c(G=0, D=1, A=2),
                       y=c(G=0, D=1, A=0))
plot(g)
```


```{r}
data("NWOGrants")
NWOGrants
```

plot it
```{r}
NWOGrants %>% 
  mutate(success=awards/applications) %>%
  ggplot(aes(x=discipline, y=success, color=gender, size=applications)) +
  geom_point() +
  theme(axis.text.x = element_text(angle=90, hjust = 1, vjust=0.5))
```

overall difference, irrespective of field
```{r}
d1 <- with(NWOGrants, list(g=ifelse(gender=="m",1,2),
                           applications=applications,
                           awards=awards))

m11.1 <- ulam(
  alist(awards ~ dbinom(applications, p),
        logit(p) <- a[g],
        a[g] ~ dnorm(0,1.5)),
  data=d1,
  chains = 4,
  cores = 4)
```

this is on the logit scale
```{r}
precis(m11.1, depth=2)
```

look at differences in award rate
```{r}
post <- extract.samples(m11.1) 

# relative scale
precis(data.frame(rel_dif=exp(post$a[,2]-post$a[,1])))

#absolute scale
precis(data.frame(prob_dif=inv_logit(post$a[,2])-inv_logit(post$a[,1])))
```
Women are 82% as likely to receive an award, translating to a reduced success rate of 3% 



## now fit a model that has a separate probability for each discipline

```{r}
d2 <- with(NWOGrants, list(g=ifelse(gender=="m",1,2),
                           applications=applications,
                           awards=awards, 
                           discipline=rep(1:9, each=2)))

m11.2 <- ulam(
  alist(awards ~ dbinom(applications, p),
        logit(p) <- a[g] + b[discipline],
        a[g] ~ dnorm(0,1.5),
        b[discipline] ~ dnorm(0,1.5)),
  data=d2,
  iter=2000,
  chains = 4,
  cores = 4)
```

```{r}
pairs(m11.2)
precis(m11.2, depth=2)
```

so much correlation.  Try not indexing gender. I this parameterization each discipline coefficient will be the rate for males in that discipline and then the gender coefficient will be the difference for females.

```{r}
d3 <- with(NWOGrants, list(g=ifelse(gender=="m",0,1),
                           applications=applications,
                           awards=awards, 
                           discipline=rep(1:9, each=2)))

m11.3 <- ulam(
  alist(awards ~ dbinom(applications, p),
        logit(p) <-  a[discipline] + b_female*g,
        a[discipline] ~ dnorm(0,1.5),
        b_female ~dnorm(0,1.5)),
  data=d3,
  iter=2000,
  chains = 4,
  cores = 4)
```


```{r}
precis(m11.3, depth=2)
pairs(m11.3)
```
This looks much better.

look at differences in award rate.

On relative scale
```{r}
post <- extract.samples(m11.3, pars="b_female") 

# relative  and absolute scale
precis(list(rel_female=exp(post$b_female)))
```

So women are 86% as likely to get an award, but the 89% condidence intervals cross 1

For the absolute scale I think it will probably be easier to use link

```{r}
# the difference between men and women will be the same for all disciplines using this model, so just get one of them.
newdat <- data.frame(g=0:1, 
                     discipline=1)
newdat
pred <- link(m11.3, data = newdat)
head(pred)
```

```{r}
precis(list(abs_female=pred[,2] - pred[,1]))
```

Women do 3% worse when accounting for overall differences in award rate between departments, although confidence interval touches 0

Can I do this from posterior directly?

```{r}
post <- extract.samples(m11.3)
str(post)
# again I should just be able to look at one discipline
precis(list(abd_female=inv_logit(post$a[,1]) -inv_logit(post$a[,1]-post$b_female)))
```

Overall I do see a reduction in award rates to women.  When we consider discipline than the signficance of this drops, but I wonder if there is still something going on...

interaction?

```{r}


m11.4 <- ulam(
  alist(awards ~ dbinom(applications, p),
        logit(p) <-  a[discipline] + b_female*g + inter[discipline]*g,
        a[discipline] ~ dnorm(0,1.5),
        b_female ~dnorm(0,1.5),
        inter[discipline] ~ dnorm(0,.5)),
  data=d3,
  iter=2000,
  chains = 4,
  cores = 4)
```

```{r}
precis(m11.4, depth=2)
```

```{r}
plot(precis(m11.4, depth=2))
```


## 11.2

_2. Suppose that the NWO Grants sample has an unobserved confound that influences both choice of discipline and the probability of an award. One example of such a confound could be the career stage of each applicant. Suppose that in some disciplines, junior scholars apply for most of the grants. In other disciplines, scholars from all career stages compete. As a result, career stage influences discipline as well as the probability of being awarded a grant. Add these influences to your DAG from Problem 1. What happens now when you condition on discipline? Does it provide an un-confounded estimate of the direct path from gender to an award? Why or why not? Justify your answer with the back-door criterion. Hint: This is structurally a lot like the grandparents-parentschildren-neighborhoods example from a previous week. If you have trouble thinking this though, try simulating fake data, assuming your DAG is true. Then analyze it using the model from Problem 1. What do you conclude? Is it possible for gender to have a real direct causal influence but for a regression conditioning on both gender and discipline to suggest zero influence?_


```{r}
g <- dagitty("dag{
  G -> A;
  G -> D;
  D -> A;
  C -> A;
  C -> D;
}")
coordinates(g) <- list(x=c(G=0, D=1, A=2, C=2),
                       y=c(G=0, D=1, A=0, C=1))
plot(g)
```

So if this is the DAG, the regression model from 1 closes the D->A, but leaves a back door from D through C to A?

# Chapter 11.2 problems

## 10E4
_Why do Poisson regressions sometimes require the use of an offset? Provide an example._

If the data have been collected (events counted) over different sampling times/ distances (exposures) we need an offset to account for this.  For example, we want to compare transposition events across the genome.  One group counted number of transposons per 100KB and the other per 1MB.

## 10M2
_If a coefficient in a Poisson regression has value 1.7, what does this imply about the change in the outcome?_

For every unit change in the predictor there will be a 5.47-fold (e^1.7) increase  in the number of counts 

```{r}
exp(1)
exp(1+1.7)
exp(1+1.7+1.7)
exp(1)^1.7
```

## 10M3
_Explain why the logit link is appropriate for a binomial generalized linear model._

We need to transform the linear model to return a value between 0 and 1 (i.e. the probability scale)

## 10M4
_Explain why the log link is appropriate for a Poisson generalized linear model._

This keeps the outcome variable on a positive scale, required for count data.  You cannot have negative counts.  

(OK but this is true for so much stuff that we model with linear models and Gaussian distributions...)

## 10H4
## Week6 PDF # 3 