---
title: "Chapter8"
author: "Julin N Maloof"
date: "7/30/2019"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rethinking)
library(tidyverse)
```

## 7E1
_For each of the causal relationships below, name a hypothetical third variable that would lead to an interaction effect._

_(1) Bread dough rises because of yeast_
Temperature

_(2) Education leads to higher income._
Job? Race? Gender?

_(3) Gasoline makes a car go._
Key? Spark?

## 7M1

## 7M2

_Recall the tulips example from the chapter. Suppose another set of treatments adjusted the temperature in the greenhouse over two levels: cold and hot. The data in the chapter were collected at the cold temperature. You find none of the plants grown under the hot temperature developed any blooms at all, regardless of the water and shade levels. Can you explain this result in terms of interactions between water, shade, and temperature?_

I am thinking that the way I would do this would be to remove the main effects of water and shade, and have W X temp, S X temp, and W X S X temp terms

## 7M2

_7M2. Can you invent a regression equation that would make the bloom size zero, whenever the temperature is hot?_

$$
bloom \sim \alpha_{[T]} + \beta_{WT} * W * T + \beta_{ST}*S*T + \beta_{WST}*W*S*T
$$
And code $T=1$ for cold and $T=0$ for hot

## 7M3

_In parts of North America, ravens depend upon wolves for their food. This is because ravens are carnivorous but cannot usually kill or open carcasses of prey. Wolves however can and do kill and tear open animals, and they tolerate ravens co-feeding at their kills. This species relationship is generally described as a “species interaction.” Can you invent a hypothetical set of data on raven population size in which this relationship would manifest as a statistical interaction? Do you think the biological interaction could be linear? Why or why not?_

Ravens ~ Wolves + Prey + Wolves:Prey

## 7H1
_Return to the data(tulips) example in the chapter. Now include the bed variable as a pre- dictor in the interaction model. Don’t interact bed with the other predictors; just include it as a main effect. Note that bed is categorical. So to use it properly, you will need to either construct dummy variables or rather an index variable, as explained in Chapter ??._

```{r}
library(tidyverse)
library(rethinking)
```

```{r}
data(tulips)
tulips
d <- tulips
```

Creat the index for bed
```{r}
d$bed_i <- as.numeric(d$bed)
d
```

normalize and center
```{r}
## R code 8.20
d$blooms_std <- d$blooms / max(d$blooms)
d$water_cent <- d$water - mean(d$water)
d$shade_cent <- d$shade - mean(d$shade)
```

run original model
```{r}
## R code 8.24
m8.7 <- quap(
  alist(
    blooms_std ~ dnorm( mu , sigma ) ,
    mu <- a + bw*water_cent + bs*shade_cent + bws*water_cent*shade_cent ,
    a ~ dnorm( 0.5 , 0.25 ) ,
    bw ~ dnorm( 0 , 0.25 ) ,
    bs ~ dnorm( 0 , 0.25 ) ,
    bws ~ dnorm( 0 , 0.25 ) ,
    sigma ~ dexp( 1 )
  ) ,
  data=d )
```

modified model with bed
```{r}
m8.7bed <- quap(
  alist(
    blooms_std ~ dnorm( mu , sigma ) ,
    mu <- a[bed_i] + bw*water_cent + bs*shade_cent + bws*water_cent*shade_cent ,
    a[bed_i] ~ dnorm( 0.5 , 0.25 ) ,
    bw ~ dnorm( 0 , 0.25 ) ,
    bs ~ dnorm( 0 , 0.25 ) ,
    bws ~ dnorm( 0 , 0.25 ) ,
    sigma ~ dexp( 1 )
  ) ,
  data=d )
```

take a look
```{r}
precis(m8.7)
```

```{r}
precis(m8.7bed, depth = 2)
```

```{r}
plot(coeftab(m8.7, m8.7bed))
```

we see that the confidence intervals on the beta coefficients have slightly shrunk, and that sigma is slightly smaller.

## 7H2
_Use WAIC to compare the model from 7H1 to a model that omits bed. What do you infer from this comparison? Can you reconcile the WAIC results with the posterior distribution of the bed coefficients?_

```{r}
compare(m8.7, m8.7bed)
```

The very small gains in parameter estimates are offset by the additional parameters. Similar models, no reason to prefer the one including bed.

## 7H3

_Use the tomato.csv (attached) data set and evaluate whether hypocotyl length ("hyp") is affected by shade ("trt"), species ("species") and their interaction._

```{r}
d <- read_csv("Tomato.csv") %>%
  select(hyp, trt, species) %>%
  na.omit()
d
```

Make a plot
```{r}
d %>%
  group_by(species,trt) %>%
  summarize(mean=mean(hyp), 
            sem=sd(hyp)/sqrt(n()), 
            ymax=mean+sem,
            ymin=mean-sem) %>%
  ggplot(aes(x=species, y=mean, ymax=ymax, ymin=ymin, fill=trt)) +
  geom_col(position = "dodge") +
  geom_errorbar(position = position_dodge(width=.9), width=.5)
```

make indices for the factors:
```{r}
d <- d %>%
  mutate(species_i = as.numeric(as.factor(species)),
         trt_i = as.numeric(as.factor(trt))-1)
```


fit non interaction model
```{r}
m1 <- quap(flist = alist(
  hyp ~ dnorm(mu, sigma),
  mu <- a[species_i] + b*trt_i, # one beta coefficient
  a[species_i] ~ dnorm(25, 5),
  b ~ dnorm(0, 5),
  sigma ~ dexp(1)),
  data=d, start=list(b=0, sigma=3))
```

check the priors
```{r}
prior <- extract.prior(m1)
str(prior)
d2 <- expand.grid(species_i=1:5, trt_i=0:1)
prior.pred <- link(m1, post=prior, data=d2)
colnames(prior.pred) <- str_c("species", d2$species_i, "_trt", d2$trt_i)
prior.pred %>% as_tibble() %>%
  gather() %>%
  ggplot(aes(x=key, y=value)) +
  geom_violin() +
  theme(axis.text.x = element_text(angle=90))
```


```{r}
precis(m1, depth=2)
```

now the interaction model:
```{r}
m2 <- quap(flist = alist(
  hyp ~ dnorm(mu, sigma),
  mu <- a[species_i] + b_int[species_i]*trt_i, # a beta coefficent for each species
  a[species_i] ~ dnorm(25, 5),
  b ~ dnorm(0, 5),
  b_int[species_i] ~ dnorm(0, 1),
  sigma ~ dexp(1)),
  data=d, start=list(b=0, sigma=3), control=list(maxit=500))
```

```{r}
precis(m2, depth=2)
```

```{r}
compare(m1,m2)
```

```{r}
m3 <- lm(hyp ~ species*trt, data=d)
summary(m3)
```


## quick dummy variable creation

```{r}
dummy <- tibble(ID=1:100, trt=sample(LETTERS[1:4], size = 100, replace=TRUE))
dummy %>% spread(key=trt,value=trt) %>%
  mutate_at(.vars=vars(-ID), ~ ifelse(is.na(.), 0, 1))
```

## 7H4
_The values in data(nettle) are data on language diversity in 74 nations. The meaning of each column is given below._
(1) country: Name of the country  
(2) num.lang: Number of recognized languages spoken  
(3) area: Area in square kilometers  
(4) k.pop: Population, in thousands  
(5) num.stations: Number of weather stations that provided data for the next two columns  
(6) mean.growing.season: Average length of growing season, in months  
(7) sd.growing.season: Standard deviation of length of growing season, in months  

_Use these data to evaluate the hypothesis that language diversity is partly a product of food security. The notion is that, in productive ecologies, people don’t need large social networks to buffer them against risk of food shortfalls. This means ethnic groups can be smaller and more self-sufficient, leading to more languages per capita. In contrast, in a poor ecology, there is more subsistence risk, and so human societies have adapted by building larger networks of mutual obligation to provide food insurance. This in turn creates social forces that help prevent languages from diversifying._

_Specifically, you will try to model the number of languages per capita as the outcome variable:_

d$lang.per.cap <- d$num.lang / d$k.pop

_Use the logarithm of this new variable as your regression outcome. (A count model would be better here, but you’ll learn those later, in Chapter 11.)_

_This problem is open ended, allowing you to decide how you address the hypotheses and the uncertain advice the modeling provides. If you think you need to use WAIC anyplace, please do. If you think you need certain priors, argue for them. If you think you need to plot predictions in a certain way, please do. Just try to honestly evaluate the main effects of both mean.growing.season and sd.growing.season, as well as their two-way interaction, as outlined in parts (a), (b), and (c) below. If you are not sure which approach to use, try several._

_(a) Evaluate the hypothesis that language diversity, as measured by log(lang.per.cap), is pos- itively associated with the average length of the growing season, mean.growing.season. Consider log(area) in your regression(s) as a covariate (not an interaction). Interpret your results._

setup
```{r}
library(tidyverse)
library(rethinking)
data("nettle")
d <- nettle
d$lang.per.cap <- d$num.lang / d$k.pop
```


Standardize the predictors
```{r}
d$mean.growing.season.scale <- scale(d$mean.growing.season)
d$sd.growing.season.scale <- scale(d$sd.growing.season)
d$area.scale <- scale(d$area)
```


take a look at the relationships

```{r, fig.height=4, fig.width=4}
qplot(d$mean.growing.season.scale, log(d$lang.per.cap)) + geom_smooth()
qplot(d$sd.growing.season.scale, log(d$lang.per.cap)) + geom_smooth()
qplot(d$mean.growing.season.scale, d$sd.growing.season.scale) + geom_smooth()

```

```{r}
m7h4a.1 <- quap(flist = alist(
  log(d$lang.per.cap) ~ dnorm(mu, sigma),
  mu <- a + bMGS*mean.growing.season.scale,
  a ~ dnorm(-3.8,2), # encompasses the range of the data
  bMGS ~ dnorm(0, 1), # allows up to a slope of 2, which would allow 8 units of change in log(lang.per.cap)
  sigma ~ dexp(1)),
  data=d)
```


check the priors
```{r}
prior <- extract.prior(m7h4a.1)
prior.pred <- link(m7h4a.1, 
                   post=prior,
                   n=100,
                   data=list(mean.growing.season.scale=range(d$mean.growing.season.scale)))

colnames(prior.pred) <- round(range(d$mean.growing.season.scale),2)

prior.pred %>%
  as_tibble() %>%
  mutate(id=1:nrow(.)) %>%
  gather(key="mean.growing.season", value="log.lang.per.cap", -id) %>%
  mutate(mean.growing.season = as.numeric(mean.growing.season)) %>%
  ggplot(aes(x=mean.growing.season, y=log.lang.per.cap, group=id)) +
  geom_line(alpha=.05)
```

A little broad, but not ridiculous.

now include area

```{r}
m7h4a.2 <- quap(flist = alist(
  log(d$lang.per.cap) ~ dnorm(mu, sigma),
  mu <- a + bMGS*mean.growing.season.scale + bA*area.scale,
  a ~ dnorm(-3.8,2), # encompasses the range of the data
  bMGS ~ dnorm(0, 1), # allows up to a slope of 2, which would allow 8 units of change in log(lang.per.cap),
  bA ~ dnorm(0,1),
  sigma ~ dexp(1)),
  data=d)
```

```{r}
precis(m7h4a.1)
precis(m7h4a.2)
plot(coeftab(m7h4a.1, m7h4a.2))
```

```{r}
#compare(m7h4a.1, m7h4a.2) #getting an error
```

Although I can't get compare to work, it does not seem that including Area improves model fit, I would stay with the simpler model.  The model supports the hypothesis that longer growing seasons are associated with an increase in the number of languages per capita.

_(b) Now evaluate the hypothesis that language diversity is negatively associated with the standard deviation of length of growing season, sd.growing.season. This hypothesis follows from uncertainty in harvest favoring social insurance through larger social networks and therefore fewer languages. Again, consider log(area) as a covariate (not an interaction). Interpret your results._

```{r}
m7h4b.1 <- quap(flist = alist(
  log(d$lang.per.cap) ~ dnorm(mu, sigma),
  mu <- a + bSDGS*sd.growing.season.scale,
  a ~ dnorm(-3.8,2), # encompasses the range of the data
  bSDGS ~ dnorm(0, 1), # allows up to a slope of 2, which would allow 8 units of change in log(lang.per.cap)
  sigma ~ dexp(1)),
  data=d)
```

now include area

```{r}
m7h4b.2 <- quap(flist = alist(
  log(d$lang.per.cap) ~ dnorm(mu, sigma),
  mu <- a + bSDGS*sd.growing.season.scale + bA*area.scale,
  a ~ dnorm(-3.8,2), # encompasses the range of the data
  bSDGS ~ dnorm(0, 1), # allows up to a slope of 2, which would allow 8 units of change in log(lang.per.cap),
  bA ~ dnorm(0, 1),
  sigma ~ dexp(1)),
  data=d)
```

```{r}
precis(m7h4b.1)
precis(m7h4b.2)
plot(coeftab(m7h4b.1, m7h4b.2))
```

```{r}
#compare(m7h4b.1, m7h4b.2) #getting an error
```

Both models support that standard deviation in growing season is negatively associated with the number of languages per capita.  There does not seem to be much area effect.

_(c) Finally, evaluate the hypothesis that mean.growing.season and sd.growing.season interact to synergistically reduce language diversity. The idea is that, in nations with longer average growing seasons, high variance makes storage and redistribution even more important than it would be otherwise. That way, people can cooperate to preserve and protect windfalls to be used during the droughts. These forces in turn may lead to greater social integration and fewer languages._

```{r}
m7h4c.1 <- quap(flist = alist(
  log(d$lang.per.cap) ~ dnorm(mu, sigma),
  mu <- a + 
    bMGS*mean.growing.season.scale + 
    bSDGS*sd.growing.season.scale +
    bInt*mean.growing.season.scale*sd.growing.season.scale,
  a ~ dnorm(-3.8,2), # encompasses the range of the data
  bSDGS ~ dnorm(0, 1), # allows up to a slope of 2, which would allow 8 units of change in log(lang.per.cap)
  bMGS ~ dnorm(0, 1),
  bInt ~ dnorm(0, 1),
  sigma ~ dexp(1)),
  data=d)
```

now include area

```{r}
m7h4c.2 <- quap(flist = alist(
  log(d$lang.per.cap) ~ dnorm(mu, sigma),
  mu <- a + 
    bA*area.scale + 
    bMGS*mean.growing.season.scale + 
    bSDGS*sd.growing.season.scale +
    bInt*mean.growing.season.scale*sd.growing.season.scale,
  a ~ dnorm(-3.8,2), # encompasses the range of the data
  bSDGS ~ dnorm(0, 1), # allows up to a slope of 2, which would allow 8 units of change in log(lang.per.cap)
  bMGS ~ dnorm(0, 1),
  bInt ~ dnorm(0, 1),
  bA ~ dnorm(0,1),
  sigma ~ dexp(1)),
  data=d)
```

```{r}
precis(m7h4c.1)
precis(m7h4c.2)
plot(coeftab(m7h4c.1, m7h4c.2))
```

still no area effect.  some evidence for negative interacton term. 

negative interaction term means that if there is a long growing season and high standard deviation, then there are fewer languages than you would otherwise expect.

compare coeftab for the three no-area models.

```{r}
plot(coeftab(m7h4a.1, m7h4b.1, m7h4c.1))
```

make some counterfactual plots

First I create a tibble of data frames, one for each level of sd.growing.season.scale that I want to look at
```{r}
counterfact.data <-
  expand.grid(sd.growing.season.scale=c(-1.5, 0, 1.5),
              mean.growing.season.scale=c(-2,2)) %>%
  as_tibble() %>%
  mutate(key=sd.growing.season.scale) %>%
  group_by(key) %>%
  nest()

counterfact.data

counterfact.data$data[[1]]
```

Now I run the link function on each of these data sets
```{r}
counterfact.predictions <- counterfact.data %>%
  mutate(pred = map(data, ~ as_tibble(link(m7h4c.1, n=20, data=.))))
counterfact.predictions
```

And finally, make a plot

```{r}
counterfact.predictions %>% unnest(pred) %>%
  rename(sd.growing.season.scale=key,
         `-2` = V1,
         `2` = V2) %>%
  mutate(id=1:nrow(.)) %>%
  gather(key="mean.growing.season.scale", 
         value="log.lang.per.cap", 
         -sd.growing.season.scale,-id) %>%
  mutate(mean.growing.season.scale=as.numeric(mean.growing.season.scale)) %>%
  
  ggplot(aes(x=mean.growing.season.scale, y=log.lang.per.cap, group=id)) +
  facet_wrap(~ sd.growing.season.scale) +
  geom_line(alpha=.2)

```

So, the model predicts that at low SD of growing season there is a strong positive relationship between growing season length and languages per cap, and this relationshp diminishes with increased SD.
