---
title: "Chapter4"
author: "Julin N Maloof"
date: "4/23/2019"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(rethinking)
```

4E1-4E5; 4M1, 4M2, 4M3

## 4E1. 
_In the model definition below, which line is the likelihood?_

$$
y_i \sim Normal(\mu, \sigma)
$$

## 4E2. 
_In the model definition just above, how many parameters are in the posterior distribution?_

Two, $\mu, \sigma$

## 4E3. 
_Using the model definition above, write down the appropriate form of Bayes’ theorem that includes the proper likelihood and priors._


## 4E4. 
_In the model definition below, which line is the linear model?_


$$
\mu_i = \alpha + \beta x_i
$$

## 4E5. 
_In the model definition just above, how many parameters are in the posterior distribution?_

Three, $\alpha, \beta, \sigma$


## 4M1. 
_For the model definition below, simulate observed heights from the prior (not the posterior)._

```{r}
prior_h <- rnorm(n = 10000,
                 mean = rnorm(n = 10000, mean = 0, sd = 10),
                 sd = runif(n = 10000, min = 0, max = 10))

  qplot(prior_h, geom="density", fill=I("skyblue"))
```


## 4M2. 
_Translate the model just above into a quap formula._

```{r}
f <- alist(
  height ~ dnorm(mu, sigma),
  mu ~ dnorm(0, 10),
  sigma ~ dunif(0, 10))
```


## 4M3. 
_Translate the quap model formula below into a mathematical model definition._

```{r}
flist <- alist(
    y ~ dnorm( mu , sigma ),
    mu <- a + b*x,
    a ~ dnorm( 0 , 50 ),
    b ~ dunif( 0 , 10 ),
    sigma ~ dunif( 0 , 50 )
)
```

$$
y_i \sim Normal (\mu_i, \sigma) \\
mu_i = \alpha + \beta x_i \\
\alpha \sim Normal(0, 50) \\
\beta \sim Uniform(0, 10) \\
\sigma \sim Uniform(0, 50)
$$

# Week 2 problems

4M4, 4M5, 4M6, 4H1, 4H2

## 4M4
_A sample of students is measured for height each year for 3 years. After the third year, you want to fit a linear regression predicting height using year as a predictor. Write down the mathematical model definition for this regression, using any variable names and priors you choose. Be prepared to defend your choice of priors._

$$
y_i \sim Normal (\mu_i, \sigma) \\
mu_i = \alpha + \beta*year_i \\
\alpha \sim Normal(100, 50) \\
\beta \sim Log-Normal(0, 1) \\
\sigma \sim Uniform(0, 50)\\
$$
Really we should fit each student as a random effect, but we aren't there yet.

Assume height measured in cm.  We don't know the age of the students but I am assuming elementary age.  100cm with a broad SD.

Choose log-normal for Beta because students should shrink with time.

how much can a student grow in a year? 10cm?

```{r}
betas <- seq(0,20,by=.1)
plot(betas,dlnorm(betas,2,1),type="l")
```

not entirely happy with this...

## 4M5
_Now suppose I tell you that the average height in the first year was 120 cm and that every student got taller each year. Does this information lead you to change your choice of priors? How?_

Not so different, but I would go with:

$$
y_i \sim Normal (\mu_i, \sigma) \\
mu_i = \alpha + \beta*year_i \\
\alpha \sim Normal(120, 50) \\
\beta \sim Log-Normal(0, 1) \\
\sigma \sim Uniform(0, 50)\\
$$

## 4M6
_Now suppose I tell you that the variance among heights for students of the same age is never more than 64cm. How does this lead you to revise your priors?_

$$
y_i \sim Normal (\mu_i, \sigma) \\
mu_i = \alpha + \beta*year_i \\
\alpha \sim Normal(120, 50) \\
\beta \sim Log-Normal(0, 1) \\
\sigma \sim Uniform(0, 8)\\
$$

## 4H1
 _The weights listed below were recorded in the !Kung census, but heights were not recorded for these individuals. Provide predicted heights and 89% intervals (either HPDI or PI) for each of these individuals. That is, fill in the table below, using model-based predictions._

```{r}
weights <- c(46.95,43.72,64.78,32.59,54.63)
```

These all seem to be adult weights, so I am going to use the simple regression model on adults

```{r}
#4.42
data(Howell1)
d <- Howell1
d2 <- d[ d$age >= 18 , ]

# define the average weight, x-bar
xbar <- mean(d2$weight)

# fit model
m4.3 <- quap(
    alist(
        height ~ dnorm( mu , sigma ) ,
        mu <- a + b*( weight - xbar ) ,
        a ~ dnorm( 178 , 20 ) ,
        b ~ dlnorm( 0 , 1 ) ,
        sigma ~ dunif( 0 , 50 )
    ) ,
    data=d2 )

```


```{r}
sim.heights <- sim(fit = m4.3,
                   data = data.frame(weight=weights),
                   n=1e4)
dim(sim.heights)
head(sim.heights)
```

```{r}
results <- tibble(
  individual=1:5,
  weight=weights,
  predicted.height=apply(sim.heights,2,mean)
  )
results <- cbind(results,t(apply(sim.heights, 2, HPDI)))
knitr::kable(results,digits=2)
```

## 4H2. 
_Select out all the rows in the Howell1 data with ages below 18 years of age. If you do it right, you should end up with a new data frame with 192 rows in it._

```{r}
d3 <- Howell1 %>% as_tibble() %>% filter(age < 18)
d3
```


_(a) Fit a linear regression to these data, using quap. Present and interpret the estimates. For every 10 units of increase in weight, how much taller does the model predict a child gets?_

first let's take a look
```{r}
qplot(weight, height, data = d3)
```

so not linear here, but based on problem I think we fit a simple regression model

```{r}
# define the average weight, x-bar
xbar <- mean(d3$weight)

# fit model
m4H2 <- quap(
    alist(
        height ~ dnorm( mu , sigma ) ,
        mu <- a + b*( weight - xbar ) ,
        a ~ dnorm( 100 , 20 ) ,
        b ~ dlnorm( 0 , 1 ) ,
        sigma ~ dunif( 0 , 50 )
    ) ,
    data=d3 )
```

```{r}
precis(m4H2)
```

For every 10 units increase in weight, the model predicts an increase of 27.2 height

_(b) Plot the raw data, with height on the vertical axis and weight on the horizontal axis. Super- impose the MAP regression line and 89% HPDI for the mean. Also superimpose the 89% HPDI for predicted heights._

```{r}
post <- extract.samples(m4H2)
weight.seq <- seq.int(from=min(d3$weight)-1, to=max(d3$weight)+1, by=1)
mu <- link(m4H2, data=data.frame(weight=weight.seq))
mu.mean <- apply( mu , 2 , mean )
mu.HPDI <- apply( mu , 2 , HPDI , prob=0.89 )
sim <- sim(m4H2, data=data.frame(weight=weight.seq))
sim.HPDI <- apply(sim, 2, HPDI, prob=0.89)
```

```{r}
plot( height ~ weight , data=d3 , col=rangi2 )
# plot the MAP line, aka the mean mu for each weight
lines( weight.seq , mu.mean )
shade( sim.HPDI , weight.seq, col = col.alpha("black",
                 0.15 ))
# plot a shaded region for 89% HPDI
shade( mu.HPDI , weight.seq, col = col.alpha("black",
                 0.30) )

```

try to plot with ggplot...
```{r}
plot.data <- as_tibble(cbind(weight.seq, mu.mean, t(mu.HPDI), t(sim.HPDI))) %>%
  rename(weight=weight.seq, mu.89l=`|0.89`, mu.89u=`0.89|`, sim.89l=V5, sim.89u=V6)
plot.data
```

```{r}
plot.data %>%
  ggplot(aes(x=weight)) +
  geom_line(aes(y=mu.mean)) +
  geom_ribbon(aes(ymin=sim.89l, ymax=sim.89u), fill="gray80") +
  geom_ribbon(aes(ymin=mu.89l, ymax=mu.89u), fill="gray60") +
  geom_point(aes(y=height), data=d3,color="blue", alpha=.3) +
  ylab("height")
```


_(c) What aspects of the model fit concern you? Describe the kinds of assumptions you would change, if any, to improve the model. You don’t have to write any new code. Just explain what the model appears to be doing a bad job of, and what you hypothesize would be a better model._

The model is trying to fit a straight line to curved data. This results in a poor fit, especially at the ends.  I would fit either a polynomial or spline regression to try to correct this.


            

# Code from Book

```{r}

## R code 4.1
pos <- replicate( 1000 , sum( runif(16,-1,1) ) )

## R code 4.2
prod( 1 + runif(12,0,0.1) )

## R code 4.3
growth <- replicate( 10000 , prod( 1 + runif(12,0,0.1) ) )
dens( growth , norm.comp=TRUE )

## R code 4.4
big <- replicate( 10000 , prod( 1 + runif(12,0,0.5) ) )
small <- replicate( 10000 , prod( 1 + runif(12,0,0.01) ) )

## R code 4.5
log.big <- replicate( 10000 , log(prod(1 + runif(12,0,0.5))) )

## R code 4.6
w <- 6; n <- 9;
p_grid <- seq(from=0,to=1,length.out=100)
posterior <- dbinom(w,n,p_grid)*dunif(p_grid,0,1)
posterior <- posterior/sum(posterior)

## R code 4.7
library(rethinking)
data(Howell1)
d <- Howell1

## R code 4.8
str( d )

## R code 4.9
precis( d )

## R code 4.10
d$height

## R code 4.11
d2 <- d[ d$age >= 18 , ]

## R code 4.12
curve( dnorm( x , 178 , 20 ) , from=100 , to=250 )

## R code 4.13
curve( dunif( x , 0 , 50 ) , from=-10 , to=60 )

## R code 4.14
sample_mu <- rnorm( 1e4 , 178 , 20 )
sample_sigma <- runif( 1e4 , 0 , 50 )
prior_h <- rnorm( 1e4 , sample_mu , sample_sigma )
dens( prior_h )

## R code 4.15
sample_mu <- rnorm( 1e4 , 178 , 100 )
prior_h <- rnorm( 1e4 , sample_mu , sample_sigma )
dens( prior_h )

## R code 4.16
mu.list <- seq( from=140, to=160 , length.out=200 )
sigma.list <- seq( from=4 , to=9 , length.out=200 )
post <- expand.grid( mu=mu.list , sigma=sigma.list )
post$LL <- sapply( 1:nrow(post) , function(i) sum( dnorm(
                d2$height ,
                mean=post$mu[i] ,
                sd=post$sigma[i] ,
                log=TRUE ) ) )
post$prod <- post$LL + dnorm( post$mu , 178 , 20 , TRUE ) +
    dunif( post$sigma , 0 , 50 , TRUE )
post$prob <- exp( post$prod - max(post$prod) )

## R code 4.17
contour_xyz( post$mu , post$sigma , post$prob )

## R code 4.18
image_xyz( post$mu , post$sigma , post$prob )

## R code 4.19
sample.rows <- sample( 1:nrow(post) , size=1e4 , replace=TRUE ,
    prob=post$prob )
sample.mu <- post$mu[ sample.rows ]
sample.sigma <- post$sigma[ sample.rows ]

## R code 4.20
plot( sample.mu , sample.sigma , cex=0.5 , pch=16 , col=col.alpha(rangi2,0.1) )

## R code 4.21
dens( sample.mu )
dens( sample.sigma )

## R code 4.22
HPDI( sample.mu )
HPDI( sample.sigma )

## R code 4.23
d3 <- sample( d2$height , size=20 )

## R code 4.24
mu.list <- seq( from=150, to=170 , length.out=200 )
sigma.list <- seq( from=4 , to=20 , length.out=200 )
post2 <- expand.grid( mu=mu.list , sigma=sigma.list )
post2$LL <- sapply( 1:nrow(post2) , function(i)
    sum( dnorm( d3 , mean=post2$mu[i] , sd=post2$sigma[i] ,
    log=TRUE ) ) )
post2$prod <- post2$LL + dnorm( post2$mu , 178 , 20 , TRUE ) +
    dunif( post2$sigma , 0 , 50 , TRUE )
post2$prob <- exp( post2$prod - max(post2$prod) )
sample2.rows <- sample( 1:nrow(post2) , size=1e4 , replace=TRUE ,
    prob=post2$prob )
sample2.mu <- post2$mu[ sample2.rows ]
sample2.sigma <- post2$sigma[ sample2.rows ]
plot( sample2.mu , sample2.sigma , cex=0.5 ,
    col=col.alpha(rangi2,0.1) ,
    xlab="mu" , ylab="sigma" , pch=16 )

## R code 4.25
dens( sample2.sigma , norm.comp=TRUE )

## R code 4.26
library(rethinking)
data(Howell1)
d <- Howell1
d2 <- d[ d$age >= 18 , ]

## R code 4.27
flist <- alist(
    height ~ dnorm( mu , sigma ) ,
    mu ~ dnorm( 178 , 20 ) ,
    sigma ~ dunif( 0 , 50 )
)

## R code 4.28
m4.1 <- quap( flist , data=d2 )

## R code 4.29
precis( m4.1 )

## R code 4.30
start <- list(
    mu=mean(d2$height),
    sigma=sd(d2$height)
)
m4.1 <- quap( flist , data=d2 , start=start )

## R code 4.31
m4.2 <- quap(
    alist(
        height ~ dnorm( mu , sigma ) ,
        mu ~ dnorm( 178 , 0.1 ) ,
        sigma ~ dunif( 0 , 50 )
    ) , data=d2 )
precis( m4.2 )

## R code 4.32
vcov( m4.1 )

## R code 4.33
diag( vcov( m4.1 ) )
cov2cor( vcov( m4.1 ) )

## R code 4.34
library(rethinking)
post <- extract.samples( m4.1 , n=1e4 )
head(post)

## R code 4.35
precis(post)

## R code 4.36
library(MASS)
post <- mvrnorm( n=1e4 , mu=coef(m4.1) , Sigma=vcov(m4.1) )

## R code 4.37
plot( d2$height ~ d2$weight )

## R code 4.38
set.seed(2971)
N <- 100                   # 100 lines
a <- rnorm( N , 178 , 20 )
b <- rnorm( N , 0 , 10 )

## R code 4.39
plot( NULL , xlim=range(d2$weight) , ylim=c(-100,400) ,
    xlab="weight" , ylab="height" )
abline( h=0 , lty=2 )
abline( h=272 , lty=1 , lwd=0.5 )
mtext( "b ~ dnorm(0,10)" )
xbar <- mean(d2$weight)
for ( i in 1:N ) curve( a[i] + b[i]*(x - xbar) ,
    from=min(d2$weight) , to=max(d2$weight) , add=TRUE ,
    col=col.alpha("black",0.2) )

## R code 4.40
b <- rlnorm( 1e4 , 0 , 1 )
dens( b , xlim=c(0,5) , adj=0.1 )

## R code 4.41
set.seed(2971)
N <- 100                   # 100 lines
a <- rnorm( N , 178 , 20 )
b <- rlnorm( N , 0 , 1 )

## R code 4.42
# load data again, since it's a long way back
library(rethinking)
data(Howell1)
d <- Howell1
d2 <- d[ d$age >= 18 , ]

# define the average weight, x-bar
xbar <- mean(d2$weight)

# fit model
m4.3 <- quap(
    alist(
        height ~ dnorm( mu , sigma ) ,
        mu <- a + b*( weight - xbar ) ,
        a ~ dnorm( 178 , 20 ) ,
        b ~ dlnorm( 0 , 1 ) ,
        sigma ~ dunif( 0 , 50 )
    ) ,
    data=d2 )

## R code 4.43
m4.3b <- quap(
    alist(
        height ~ dnorm( mu , sigma ) ,
        mu <- a + exp(log_b)*( weight - xbar ),
        a ~ dnorm( 178 , 100 ) ,
        log_b ~ dnorm( 0 , 1 ) ,
        sigma ~ dunif( 0 , 50 )
    ) ,
    data=d2 )
```

```{r}
## R code 4.44
precis( m4.3 )
```


```{r}
## R code 4.45
round( vcov( m4.3 ) , 3 )
```


```{r}
## R code 4.46
plot( height ~ weight , data=d2 , col=rangi2 )
post <- extract.samples( m4.3 )
a_map <- mean(post$a)
b_map <- mean(post$b)
curve( a_map + b_map*(x - xbar) , add=TRUE )
```


```{r}
## R code 4.47
post <- extract.samples( m4.3 )
post[1:5,]
```


```{r}
## R code 4.48
N <- 10
dN <- d2[ 1:N , ]
mN <- quap(
    alist(
        height ~ dnorm( mu , sigma ) ,
        mu <- a + b*( weight - mean(weight) ) ,
        a ~ dnorm( 178 , 20 ) ,
        b ~ dlnorm( 0 , 1 ) ,
        sigma ~ dunif( 0 , 50 )
    ) , data=dN )

## R code 4.49
# extract 20 samples from the posterior
post <- extract.samples( mN , n=20 )

# display raw data and sample size
plot( dN$weight , dN$height ,
    xlim=range(d2$weight) , ylim=range(d2$height) ,
    col=rangi2 , xlab="weight" , ylab="height" )
mtext(concat("N = ",N))

# plot the lines, with transparency
for ( i in 1:20 )
    curve( post$a[i] + post$b[i]*(x-mean(dN$weight)) ,
        col=col.alpha("black",0.3) , add=TRUE )
```


```{r}
## R code 4.50
post <- extract.samples( m4.3 )
mu_at_50 <- post$a + post$b * ( 50 - xbar )

## R code 4.51
dens( mu_at_50 , col=rangi2 , lwd=2 , xlab="mu|weight=50" )

## R code 4.52
HPDI( mu_at_50 , prob=0.89 )

## R code 4.53
mu <- link( m4.3 )
str(mu)

## R code 4.54
# define sequence of weights to compute predictions for
# these values will be on the horizontal axis
weight.seq <- seq( from=25 , to=70 , by=1 )

# use link to compute mu
# for each sample from posterior
# and for each weight in weight.seq
mu <- link( m4.3 , data=data.frame(weight=weight.seq) )
str(mu)

## R code 4.55
# use type="n" to hide raw data
plot( height ~ weight , d2 , type="n" )

# loop over samples and plot each mu value
for ( i in 1:100 )
    points( weight.seq , mu[i,] , pch=16 , col=col.alpha(rangi2,0.1) )
```


```{r}
## R code 4.56
# summarize the distribution of mu
mu.mean <- apply( mu , 2 , mean )
mu.HPDI <- apply( mu , 2 , HPDI , prob=0.89 )

## R code 4.57
# plot raw data
# fading out points to make line and interval more visible
plot( height ~ weight , data=d2 , col=col.alpha(rangi2,0.5) )

# plot the MAP line, aka the mean mu for each weight
lines( weight.seq , mu.mean )

# plot a shaded region for 89% HPDI
shade( mu.HPDI , weight.seq )
```


```{r}
## R code 4.58
post <- extract.samples(m4.3)
mu.link <- function(weight) post$a + post$b*( weight - xbar )
weight.seq <- seq( from=25 , to=70 , by=1 )
mu <- sapply( weight.seq , mu.link )
mu.mean <- apply( mu , 2 , mean )
mu.HPDI <- apply( mu , 2 , HPDI , prob=0.89 )

## R code 4.59
sim.height <- sim( m4.3 , data=list(weight=weight.seq) )
str(sim.height)

## R code 4.60
height.PI <- apply( sim.height , 2 , PI , prob=0.89 )

## R code 4.61
# plot raw data
plot( height ~ weight , d2 , col=col.alpha(rangi2,0.5) )

# draw MAP line
lines( weight.seq , mu.mean )

# draw HPDI region for line
shade( mu.HPDI , weight.seq )

# draw PI region for simulated heights
shade( height.PI , weight.seq )
```


```{r}
## R code 4.62
sim.height <- sim( m4.3 , data=list(weight=weight.seq) , n=1e4 )
height.PI <- apply( sim.height , 2 , PI , prob=0.89 )

## R code 4.63
post <- extract.samples(m4.3)
weight.seq <- 25:70
sim.height <- sapply( weight.seq , function(weight)
    rnorm(
        n=nrow(post) ,
        mean=post$a + post$b*( weight - xbar ) ,
        sd=post$sigma ) )
height.PI <- apply( sim.height , 2 , PI , prob=0.89 )
```


```{r}
## R code 4.64
library(rethinking)
data(Howell1)
d <- Howell1
str(d)

## R code 4.65
d$weight_s <- ( d$weight - mean(d$weight) )/sd(d$weight)
d$weight_s2 <- d$weight_s^2
m4.5 <- quap(
    alist(
        height ~ dnorm( mu , sigma ) ,
        mu <- a + b1*weight_s + b2*weight_s2 ,
        a ~ dnorm( 178 , 20 ) ,
        b1 ~ dlnorm( 0 , 1 ) ,
        b2 ~ dnorm( 0 , 1 ) ,
        sigma ~ dunif( 0 , 50 )
    ) ,
    data=d )

## R code 4.66
precis( m4.5 )

```

```{r}
## R code 4.67
weight.seq <- seq( from=-2.2 , to=2 , length.out=30 )
pred_dat <- list( weight_s=weight.seq , weight_s2=weight.seq^2 )
mu <- link( m4.5 , data=pred_dat )
mu.mean <- apply( mu , 2 , mean )
mu.PI <- apply( mu , 2 , PI , prob=0.89 )
sim.height <- sim( m4.5 , data=pred_dat )
height.PI <- apply( sim.height , 2 , PI , prob=0.89 )

## R code 4.68
plot( height ~ weight_s , d , col=col.alpha(rangi2,0.5) )
lines( weight.seq , mu.mean )
shade( mu.PI , weight.seq )
shade( height.PI , weight.seq )
```

```{r}
## R code 4.69
d$weight_s3 <- d$weight_s^3
m4.6 <- quap(
    alist(
        height ~ dnorm( mu , sigma ) ,
        mu <- a + b1*weight_s + b2*weight_s2 + b3*weight_s3 ,
        a ~ dnorm( 178 , 20 ) ,
        b1 ~ dlnorm( 0 , 1 ) ,
        b2 ~ dnorm( 0 , 10 ) ,
        b3 ~ dnorm( 0 , 10 ) ,
        sigma ~ dunif( 0 , 50 )
    ) ,
    data=d )

## R code 4.70
plot( height ~ weight_s , d , col=col.alpha(rangi2,0.5) , xaxt="n" )

## R code 4.71
at <- c(-2,-1,0,1,2)
labels <- at*sd(d$weight) + mean(d$weight)
axis( side=1 , at=at , labels=round(labels,1) )

```

```{r}
## R code 4.72
library(rethinking)
data(cherry_blossoms)
d <- cherry_blossoms
precis(d)

## R code 4.73
d2 <- d[ complete.cases(d$temp) , ] # complete cases on temp
num_knots <- 15
knot_list <- quantile( d2$year , probs=seq(0,1,length.out=num_knots) )

knot_list
```

```{r}
## R code 4.74
library(splines)
B <- bs(d2$year,
    knots=knot_list[-c(1,num_knots)] ,
    degree=3 , intercept=TRUE )

## R code 4.75
plot( NULL , xlim=range(d2$year) , ylim=c(0,1) , xlab="year" , ylab="basis value" )
for ( i in 1:ncol(B) ) lines( d2$year , B[,i] )

```

```{r}
## R code 4.76
m4.7 <- quap(
    alist(
        T ~ dnorm( mu , sigma ) ,
        mu <- a + B %*% w ,
        a ~ dnorm(6,10),
        w ~ dnorm(0,1),
        sigma ~ dexp(1)
    ),
    data=list( T=d2$temp , B=B ) ,
    start=list( w=rep( 0 , ncol(B) ) ) )
```


```{r}
## R code 4.77
post <- extract.samples(m4.7)
str(post)
w <- apply( post$w , 2 , mean )
plot( NULL , xlim=range(d2$year) , ylim=c(-2,2) ,
    xlab="year" , ylab="basis * weight" )
for ( i in 1:ncol(B) ) lines( d2$year , w[i]*B[,i] )
```


```{r}
## R code 4.78
mu <- link( m4.7 )
mu_PI <- apply(mu,2,PI,0.97)
plot( d2$year , d2$temp , col=col.alpha(rangi2,0.3) , pch=16 )
shade( mu_PI , d2$year , col=col.alpha("black",0.5) )
```


```{r}
## R code 4.79
plot( height ~ weight , data=Howell1 ,
    col=col.alpha(rangi2,0.4) )

```

