---
title: "Untitled"
author: "Julin N Maloof"
date: "10/18/2019"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. 
Consider the data(Wines2012) data table. These data are expert ratings
of 20 different French and American wines by 9 different French and American judges. Your goal is to model score, the subjective rating assigned by
each judge to each wine. I recommend standardizing it.
In this first problem, consider only variation among judges and wines.
Construct index variables of judge and wine and then use these index variables to construct a linear regression model. Justify your priors. You should
end up with 9 judge parameters and 20 wine parameters. Use ulam instead of
quap to build this model, and be sure to check the chains for convergence. If
you’d rather build the model directly in Stan or PyMC3, go ahead. I just want
you to use Hamiltonian Monte Carlo instead of quadratic approximation.
How do you interpret the variation among individual judges and individual wines? Do you notice any patterns, just by plotting the differences?
Which judges gave the highest/lowest ratings? Which wines were rated worst/
best on average?

```{r}
library(rethinking)
library(tidyverse)
data("Wines2012")
head(Wines2012)
```




```{r}
Wines2012$score_std <- scale(Wines2012$score)
wlist <- with(Wines2012,
              list(judge=as.numeric(judge),
                   wine=as.numeric(wine),
                   score = score_std))
```

```{r}
m1 <- ulam(alist(
  score ~ dnorm(mu, sigma),
  mu <- a_j[judge] + a_w[wine],
  a_j[judge] ~ dnorm(0,1),
  a_w[wine] ~ dnorm(0,1),
  sigma ~ dexp(1)),
  data = wlist,
  chains = 4,
  cores = 4,
  iter=2000)
```

are the priors reasonable?
```{r}
priors <- extract.prior(m1)
str(priors)
```

each "score" would be the sum of a wine and a judge, so we could just run through some of these
```{r}
data <- data.frame(judge=sample(1:9, 1000, replace = TRUE), 
                   wine=sample(1:20, 1000, replace = TRUE))
prior.score <- link(m1,data,post=priors)
dens(prior.score)
dens(wlist$score)
```

Reasonable...

```{r}
precis(m1, depth=2)
plot(precis(m1, depth=2))
traceplot(m1, ask=FALSE)
trankplot(m1, ask=FALSE)
```

More varitaion in judges than in wines.  Only 2 wines seem to be different from mean, but maybe 4 judges.

## 2. Now consider three features of the wines and judges:
(1) flight: Whether the wine is red or white.
(2) wine.amer: Indicator variable for American wines.
(3) judge.amer: Indicator variable for American judges.
Use indicator or index variables to model the influence of these features on
the scores. Omit the individual judge and wine index variables from Problem
1. Do not include interaction effects yet. Again use ulam, justify your priors,
and be sure to check the chains. What do you conclude about the differences
among the wines and judges? Try to relate the results to the inferences in
Problem 1.

```{r}
wlist <- with(Wines2012,
              list(score = score_std,
                   red=as.numeric(flight),
                   wine_amer=wine.amer+1,
                   judge_amer=judge.amer+1
              ))

m2 <- ulam(alist(
  score ~ dnorm(mu, sigma),
  mu <- a_red[red] + a_w_amer[wine_amer] + a_j_amer[judge_amer],
  a_red[red] ~ dnorm(0,1),
  a_w_amer[wine_amer] ~ dnorm(0,1),
  a_j_amer[judge_amer] ~ dnorm(0,1),
  sigma ~ dexp(1)),
  data = wlist,
  chains = 4,
  cores = 4,
  iter=2000)

```

```{r}
precis(m2, depth=2)
plot(precis(m2, depth=2))
traceplot(m2, ask=FALSE)
trankplot(m2, ask=FALSE)
```

No effect of flight, or country of origin for wine or judge.

## 3. Now consider two-way interactions among the three features. 
You should end up with three different interaction terms in your model. These will be
easier to build, if you use indicator variables. Again use ulam, justify your
priors, and be sure to check the chains. Explain what each interaction means.
Be sure to interpret the model’s predictions on the outcome scale (mu, the
expected score), not on the scale of individual parameters. You can use link
to help with this, or just use your knowledge of the linear model instead.
What do you conclude about the features and the scores? Can you relate
the results of your model(s) to the individual judge and wine inferences from
Problem 1?

```{r}

m3 <- ulam(alist(
  score ~ dnorm(mu, sigma),
  mu <- a_red[red] + 
    a_w_amer[wine_amer] + 
    a_j_amer[judge_amer] + 
    i_red_w_amer[red]*(wine_amer-1) + # 0 for french wines, 1 for american wines
    i_red_j_amer[red]*(judge_amer-1) + # 0 for french wines, 1 for american wines
    i_w_j_amer[wine_amer]*(judge_amer-1), # 0 for french judges, 1 for american judges
  a_red[red] ~ dnorm(0,1),
  a_w_amer[wine_amer] ~ dnorm(0,1),
  a_j_amer[judge_amer] ~ dnorm(0,1),
  i_red_w_amer[red] ~ dnorm(0,.5),
  i_red_j_amer[red] ~ dnorm(0,.5),
  i_w_j_amer[wine_amer] ~ dnorm(0,.5),
  sigma ~ dexp(1)),
  data = wlist,
  chains = 4,
  cores = 4,
  iter=2000)
```

```{r}
precis(m3, depth=2)
plot(precis(m3, depth=2))
traceplot(m3, ask=FALSE)
trankplot(m3, ask=FALSE)
```

I don't see anything inersting here:

* a_red: score of white or red wines
* a_w_amer: score of french or american wines
* a_j_amer: score of french or american judges
* i_red_w_amer: interaction effect of red on french vs american wines (the amount that a white or red score shifts when a  wine is from America 
* i_red_j_amer: interaction effect of red on french vs american wines (the amount that a white or red score shifts when a  judge is from America )
* i_w_j_amer: interaction effect of wine and judge country.  The amount that the score shifts when an american or french wine is judged by an american.


