---
title: "Chapter3"
author: "Julin N Maloof"
date: "4/18/2019"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(rethinking)
```

## Easy

```{r}
p_grid <- seq( from=0 , to=1 , length.out=1000 )
prior <- rep( 1 , 1000 )
likelihood <- dbinom( 6 , size=9 , prob=p_grid )
posterior <- likelihood * prior
posterior <- posterior / sum(posterior)
set.seed(100)
samples <- sample( p_grid , prob=posterior , size=1e4 , replace=TRUE )
```


```{r}
hist(samples)
```



### 3E1. 
_How much posterior probability lies below p = 0.2?_

```{r}
sum(samples < .2) / length(samples)
```

### 3E2. 
_How much posterior probability lies above p = 0.8?_

```{r}
sum(samples > 0.8) / length(samples)
```

### 3E3. 
_How much posterior probability lies between p = 0.2 and p = 0.8?_

```{r}
sum(samples >0.2 & samples < 0.8) / length(samples)
```

### 3E4. 
_20% of the posterior probability lies below which value of p?_

```{r}
quantile(samples, 0.2)
```


### 3E5. 
_20% of the posterior probability lies above which value of p?_

```{r}
quantile(samples, 0.8)
```


### 3E6. 
_Which values of p contain the narrowest interval equal to 66% of the posterior probability?_

```{r}
HPDI(samples, 0.66)
```


### 3E7. 
_Which values of p contain 66% of the posterior probability, assuming equal posterior probability both below and above the interval?_

```{r}
PI(samples, 0.66)
```

## Medium

### 3M1. 
_Suppose the globe tossing data had turned out to be 8 water in 15 tosses. Construct the posterior distribution, using grid approximation. Use the same flat prior as before._

```{r}
p_grid <- seq(0,1, length.out = 1000)
prior <- rep(1,length(p_grid))
likelihood <- dbinom(8, 15, prob = p_grid)
plot(likelihood)
unstd.post <- prior*likelihood
post <- unstd.post / sum(unstd.post)
plot(post)
```


### 3M2. 
_Draw 10,000 samples from the grid approximation from above. Then use the samples to calculate the 90% HPDI for p._

```{r}
samples <- sample(p_grid, size = 10000, replace = TRUE, prob = post)
hist(samples)
HPDI(samples, 0.9)
```


### 3M3. 
_Construct a posterior predictive check for this model and data. This means simulate the distribution of samples, averaging over the posterior uncertainty in p. What is the probability of observing 8 water in 15 tosses?_

```{r}
predictions <- rbinom(10000, 15, prob = samples)
sum(predictions==8) / length(predictions)
```


### 3M4. 
_Using the posterior distribution constructed from the new (8/15) data, now calculate the probability of observing 6 water in 9 tosses._

```{r}
predictions9 <- rbinom(10000, 9, prob = samples)
sum(predictions9==6) / length(predictions9)

```


### 3M5.
_Start over at 3M1, but now use a prior that is zero below p = 0:5 and a constant above p = 0:5. This corresponds to prior information that a majority of the Earth’s surface is water. Repeat each problem above and compare the inferences. What difference does the better prior make? If it helps,
compare inferences (using both priors) to the true value p = 0:7._

```{r}
p_grid <- seq(0,1, length.out = 1000)
prior <- rep(1,length(p_grid)) %>% ifelse(p_grid < .5, 0, .)
likelihood <- dbinom(8, 15, prob = p_grid)
plot(likelihood)
unstd.post <- prior*likelihood
post <- unstd.post / sum(unstd.post)
plot(post)
```


_Draw 10,000 samples from the grid approximation from above. Then use the samples to calculate the 90% HPDI for p._

```{r}
samples2 <- sample(p_grid, size = 10000, replace = TRUE, prob = post)
hist(samples2)
HPDI(samples2, 0.9)
```

Confidence intervals are tighter
